---
title: "Using the genefilter function to filter genes from a microarray dataset"
author:
- name: Robert, Gentleman
  affiliation: Vignette Author's Institution
- name: Khadijah Amusat
  affiliation: Vignette translation from Sweave to R Markdown / HTML
date: March 09, 2023
output:
  BiocStyle: BiocStyle:: html_documents:
    number_sections: true
    toc: true
    toc_depth: 4
package: genefilter
vignette: >
  %\VignetteIndexEntry{Using the genefilter function to filter genes from a microarray}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction {#introduction .unnumbered}

The *genefilter* package can be used to filter (select) genes from a
microarray dataset according to a variety of different filtering
mechanisms. Here, we will consider the example dataset in the
`sample.ExpressionSet` example from the *Biobase* package. This
experiment has 26 samples, and there are 500 genes and 3 covariates. The
covariates are named `sex`, `type` and `score`. The first two have two
levels and the last one is continuous.

```{r}
library(Biobase) 
library(genefilter)
data(sample.ExpressionSet)
 varLabels(sample.ExpressionSet)
 table(sample.ExpressionSet$sex)
 table(sample.ExpressionSet$type)
 data(sample.ExpressionSet)
 
```

One dichotomy that can be of interest for subsequent analyses is whether
the filter is *specific* or *non-specific*. Here, specific means that we
are filtering with reference to sample metadata, for example, `type`.
For example, if we want to select genes that are differentially
expressed in the two groups defined by `type`, that is a specific
filter. If on the other hand we want to select genes that are expressed
in more than 5 samples, that is an example of a non--specific filter.

First, let us see how to perform a non--specific filter. Suppose we want
to select genes that have an expression measure above 200 in at least 5
samples. To do that we use the function `kOverA`.

There are three steps that must be performed.

1.  Create function(s) implementing the filtering criteria.

2.  Assemble it (them) into a (combined) filtering function.

3.  Apply the filtering function to the expression matrix.

```{r}
f1 <- kOverA(5, 200)
ffun <- filterfun(f1)
wh1 <- genefilter(exprs(sample.ExpressionSet), ffun)
sum(wh1)
```

Here `f1` is a function that implies our "expression measure above 200
in at least 5 samples" criterion, the function `ffun` is the filtering
function (which in this case consists of only one criterion), and we
apply it using `genefilter`. There were genes that satisfied the
criterion and passed the filter.
As an example for a specific filter, let us select genes that are
differentially expressed in the groups defined by `type`.

```{r}
f2 <- ttest(sample.ExpressionSet$type, p=0.1)
ffun <- filterfun(f1)
wh2 <- genefilter(exprs(sample.ExpressionSet), 
filterfun(f2))
sum(wh2)
```
Here, `ttest` is a function from the `genefilter`
package which provides a suitable wrapper around \texttt{t.test} from
package \textit{stats}. Now we see that there are  `r \Sexpr{sum(wh2)}`
genes that satisfy the selection criterion. 

Suppose that we want to combine the two filters. We want those genes
for which at least 5 have an expression measure over 200 *and* which also are differentially
expressed between the groups defined by `type+`.

```{r}
ffun_combined <- filterfun(f1, f2)
wh3 <- genefilter(exprs(sample.ExpressionSet), ffun_combined)
sum(wh3)
```

Now we see that there are only \Sexpr{sum(wh3)} genes  that satisfy both conditions.

FIXME: need to replace this with something else
Our last example is to select genes that are
differentially expressed in at least one of the three groups defined
by \verb+cov3+.
To do that we use an Anova filter. This filter uses an analysis of
variance appraoch (via the \verb+lm+) function to test the hypothesis
that at least one of the three group means is different from the other
two. The test is applied, then the $p$--value computed. We select
%those genes that have a low $p$--value.


Afilter <- Anova(eset$cov3)
aff <- filterfun(Afilter)
wh4 <- genefilter(exprs(eset), aff)
sum(wh4)

We see that there are 14 genes that pass this filter and that are
candidates for further exploration.

\section*{Selecting genes that appear useful for prediction}

The function \texttt{knnCV} defined below performs$k$--nearest neighbour classification
using leave--one--out cross--validation.
At the same time it aggregates the genes that were selected. The
function returns the predicted classifications as its returned
value. However, there is an additional side effect. The number of
times that each gene was used (provided it was at least one) are
recorded and stored in the environment of the aggregator \verb+Agg+.
These can subsequently be retrieved and used for other purposes.

```{r}
knnCV <- function(x, selectfun, cov, Agg, pselect = 0.01, scale=FALSE) {
   nc <- ncol(x)
   outvals <- rep(NA, nc)
   for(i in seq_len(nc)) {
      v1 <- x[,i]
      expr <- x[,-i]
      glist <- selectfun(expr, cov[-i], p=pselect)
      expr <- expr[glist,]
      if( scale ) {
        expr <- scale(expr)
        v1 <- as.vector(scale(v1[glist]))
      }
      else
         v1 <- v1[glist]
      out <- paste("iter ",i, " num genes= ", sum(glist), sep="")
      print(out)
      Aggregate(row.names(expr), Agg)
      if( length(v1) == 1)
         outvals[i] <- knn(expr, v1, cov[-i], k=5)
      else
          outvals[i] <- knn(t(expr), v1, cov[-i], k=5)
    }
    return(outvals)
}
```

```{r}
 gfun <- function(expr, cov, p=0.05) {
    f2 <- ttest(cov, p=p)
    ffun <- filterfun(f2)
    which <- genefilter(expr, ffun)
  }
```

Next we show how to use this function on the dataset
\verb+geneData+.

<<aggregate2, results=hide>>=
  library("class")

  ##scale the genes
  ##genescale is a slightly more flexible "scale"
  ##work on a subset -- for speed only
  
  geneData <- genescale(exprs(sample.ExpressionSet)[1:75,], 1)
  Agg <- new("aggregator")
  testcase <- knnCV(geneData, gfun, sample.ExpressionSet$type, Agg,pselect=0.05) 
aggregate3 <- sort(sapply(aggenv(Agg), c),
decreasing=TRUE) 

The environment `Agg` contains, for each gene, the
number of times it was selected in the cross-validation.

# Session Information {#session-information .unnumbered}

The version number of R and packages loaded for generating the vignette
were:

\<\<sessionInfo,echo=FALSE,results=tex\>\>= toLatex(sessionInfo()) @